Wealth dynamics in the presence of network structure and primitive cooperation

### 2019-07-18 manuscript/draftcssa19.tex
Minor edits & grammer fixes

\maketitle
\begin{abstract}
We study wealth accumulation dynamics in a population of heterogeneously mixed agents with a capacity for a certain primitive form of cooperation enabled by static network structures. Despite their simplicity, the stochastic dynamics generate inequalities in wealth reminiscent of real world social systems even in a fully mixed population. A simple form of cooperation is introduced and is shown to enhance the viability of agents By embedding such dynamics in a network, the impact of social structures on the origins and persistence of inequality can be teased out easily. The models developed here complement traditional modeling approaches based on grid worlds.

\end{abstract}
\section{Introduction}
In recent years, concerns and debates surrounding wealth inequality and socioeconomic mobility have been one of the few unifying issues dominating the extremely polarized public spheres of the Global North. While economic and political inequality used to be discussed in not so mainstream heterodox economics circles, the contentious discussions on the topic within mainstream economics since the publication of Piketty's book~\cite{piketty2017capital} suggests a lack of consensus about very basic foundational questions like the origins and persistence of economic inequality. Not surprisingly, traditional theories and tools of macro and micro economics are now being diagnosed for their limitations. Simultaneously, insights from related disciplines, along with novel scientific inquiry models not usually associated with traditional econometrics, are being taken more seriously. The work presented here shares this spirit by integrating substantive ideas from anthropology, economic sociology~\cite{granovetter2017society} and urban sociology~\cite{sampson2012great} and using modeling approaches from computational social science~\cite{hedstrom2018}, computational economics~\cite{tesfatsion} and analytical sociology~\cite{hedstrom2011oxford} to understand the origins of inequality in a simple model of wealth dynamics in the presence of social structures.    
% SD: citation for "anthropology," the first element of the last sentence's list? Grief and/or White?

The current work originated in our attempts to incorporate and tease out the effects of social structures in simple models of wealth dynamics in the presence of environmental stochasticity, and a simple form of resource pooling. Specifically, we are interested in studying how structure and dynamics interact to produce and sustain inequality, and influence robustness to scarcity shocks. We answer the former question using Gini coefficients; the latter question is answered using survival time type analysis. As we discuss in section 3, while Gini was found to be of limited value, survival time analysis produced more insights on both questions of inequality and robustness.

These models were inspired by our search for analogs of the grid world agent-based models (ABMs) of Friesen and Mudigonda~\cite{srimil} where foraging agents that pool their resources were shown to on-average outperform non-pooling agents. The original model used foraging to mix the population and create opportunities for interaction, and, when certain conditions were met, allowed resource sharing. Our model achieves agent interaction by postulating a static network structure which partially mixes the agents. Of primary interest is whether network effects alone can generate and sustain differences in economic outcomes of otherwise homogeneous agents.

The original model drew its inspiration from historical sociology, in Katz's influential study of middle of 19th century Hamilton, Canada~\cite{katz2013people}. Retaining this original motivation, we draw additional inspiration from economic sociology, in the work of Granovetter~\cite{granovetter2005}; urban sociology, in the work of Sampson~\cite{sampson2002} and others; and in anthropology, in the work on cooperation in small to medium scale societies~\cite{avner1994,white2011kinship}. The geographic and economic scale of the systems, the nature of social actors and time period of interest are all very different from the ones used to develop models of representative agents in macroeconomics~\cite{benhabib2018}, making the similarity between macroeconomic wealth dynamics models and our models not comparable without further justification. Elaborating on the interplay of conceptual and methodological ideas among these disciplines is beyond the scope of this article. Instead, we anchor this work in economic sociology and revisit the insights from the above disciplines in the article's concluding section.

The important role played by social structures or non-economic structures in determining economic outcomes of individuals in a society is not in doubt~\cite{granovetter2005,jackson_rev2017}. Still, in the absence of a unifying foundation for sociology and economics, the full impact of this two-way interpenetration of economic and social structures is demonstrated only on a case-by-case basis. The language of social and economic networks affords a first principle integration by simplifying the non-trivial concept of social and economic structure~\cite{martin_lee} to only dyadic relations.

As mentioned earlier, macroeconomic models are ill-suited to the study of collective phenomena at an intermediate level. Hence, alternative explanations of aggregate phenomena that match the expressiveness of economic models are required. Analytical sociology~\cite{ch1as_hdbk}, with its emphasis on explanation of collective emergent phenomena using mathematically formulated social mechanisms~\cite{ch2as_hdbk,ch11as_hdbk} anchored at the individual level, is an ideal candidate for this purpose. It is particularly powerful when combined with computational simulation, because when mathematically-formulated models reach even a modest level of complexity, they often become analytically intractable. Simulation \textit{in silico} can yield approximate results for these more complex scenarios, which supplement the exact results reached by traditional methods.

@@

\label{giniKindaSucks}
% SD: In light of recent discussions, we may want to downplay Gini and up-play the robustness of agents to exogenous shocks.
The phenomenon under scrutiny is the emergence and evolution of economic inequality in societies with non-trivial social structures that enable economic interactions and cooperation mechanisms. The model we use to answer questions surrounding this phenomenon must possess a dynamic model of wealth accumulation, a model of social structure, and a suitably useful measure of inequality. The dynamics are modeled as Brownian noise driven linear dynamical systems, here constant growth rate; the structure is modeled by networks, here random graphs; the measures of inequality used are Gini coefficients, and survival time distributions of agents in response to lack of resources. Although both the dynamical system and the network model are quite well understood, the precise interplay of structure and dynamics produces interesting emergent wealth distributions and robustness to scarcity. To the best of our knowledge, this specific combination of network structure and social dynamics has not been discussed in the computational social science (CSS) or mathematical sociology literature, and we consider this the primary contribution of this work.

@@

In the next section, we discuss the mathematical formulation of the model. In section 3, we present the analysis of our simulation experiments, summarize key findings and discuss why we chose the \texttt{julia} programming language~\cite{Julia-2017} for our implementation. In section 4, we discuss the limitations of our simple models, extensions to dynamics and networks more expressive than the ones presented here, and planned future work.

Before discussing our model in greater mathematical detail, we discuss the model qualitatively, contrasting them with more familiar modeling approaches. The model presented here has much in common with models used in social reality inspired models in statistical physics~\cite{redner2001guide}, dynamic process models in network science~\cite{newman2018networks}, and computational social science and agent-based models~\cite{abm_review}; however, our modeling philosophy is somewhere at the interface of agent-based computational economics (ACE)~\cite{tesfatsion} and analytical sociology (AS)~\cite{ch1as_hdbk,ch2as_hdbk}. In ACE, we acknowledge its aspiration to develop bottom-up models of economic systems at all scales but restrain from its enthusiastic use of complex but well-calibrated detailed models of markets and agents~\cite{tesfatsion2017}. In AS, we assent its focus on social mechanisms in explaining social phenomena, but instead rely on simplified mechanisms with few parameters~\cite{ch11as_hdbk} with the specific goal of extracting insights from \textit{stylized} models.

More specifically, from statistical physics, we borrow the dynamics: diffusion models and associated first-passage time techniques; from network science, we borrow the structural aspects: Erd\H{o}s-R\'{e}nyi network (ER) models; and from non-equlibrium statistical physics, CSS and ACE, we borrow a form of cooperation: the concept of coalescence, institution and coordination.

\section{Model}
The model of cooperative wealth accumulation constructed here is best thought of as a stochastic interacting particle system infused with economic sociological semantics. The particle evolves according to an one dimensional diffusion process with constant drift and is driven by a Brownian noise with boundary condition at the origin corresponding to particle absorption. After crossing a pre-determined threshold in state space, particles above the threshold follow a protocol and coalesce together. In an ensemble of otherwise identical particles, a given particle may coalesce with a subset of other particles. This mixing characteristic is encoded via a graph. Both the conjoined particles and individual particles die upon crossing the origin. This model of interacting diffusing particles can be provided with substantive semantics as follows.

%Might be nice to put a nice table here 
The one dimensional state space of the particle is identified with the wealth of an social actor (agent). We consider a homogeneous population where all agents are required burn their wealth at a constant specified rate in order to survive. In addition, the agents all gain wealth at a constant rate. The resource draining rate and the resource gaining rate are additive and constitute the drift of the diffusion process. The environmental contingency is modeled by a Brownian noise of a specified intensity. Upon crossing a specific wealth threshold, agents can coalesce to form a cooperative unit deciding to pool their resources and their environmental contingencies into a single unit which we call a \textit{proto-institution} (proto). The mixing characteristic of this ensemble is the social network (here ER model).

The questions of interest to us, emergence and persistence of inequality and robustness to scarcity, in the presence of social structure and uncertain environmental conditions, map onto questions about the stochastic dynamical system (SDS). Inequality can be quantified using Gini coefficients of the particle ensemble's state space. Robustness to scarcity can be measured using survival time analysis for the particles to reach the absorbing boundary at the origin, by turning off the appropriate drift terms in the dynamics.

@@

\noindent where $v$ is the wealth growth rate (the difference of the income and metabolic rate of the agent) and $D$ is the intensity of the Brownian process (white noise process) $w$. $x(t)$ is the state of the particle (wealth) at time $t$. The dynamics can be started at a any initial point $x_0 > 0$.

@@

Since the particles are independent, they are independent dynamical systems; proto formation is a higher level construct imposed on this system and is a constraint on what states are considered viable and what are not. Consider two particles $x_1$ and  $x_2$ that have crossed the threshold for coalescence $x_{thresh}$. After the coalescence event, the state (wealth) of individual particles $x_1(t)$ and  $x_2(t)$ is not important; only the aggregate wealth of the coalescent (proto) is important. As long as $x_1(t) + x_2(t) >0$, both particles survive. Effectively, the proto is a two-dimensional SDS. Since the wealth of the proto ($p_{12}$)is additive, the aggregate wealth $p_{12} = x_1(t) + x_2(t) $. The dynamical variable satisfies an SDE\footnote{The result follows from the additivity properties of white noise.}

@@

For such ensemble of particles, the primary driver of difference in paths (life histories) and wealth is the Brownian noise intensity $D$; greater the $D$. greater the diversity (and hence Gini). However, Gini is a measure that is dependent on absolute magnitude. So,  if the drift ($v$) is large, then the variation generated by
$D$ gets washed out by $v$\footnote{Preliminary investigations suggest that for simple non-network ensembles, Gini either stays close to $0$ or $1$. We suspect that this is because of the constant wealth growth rate used in our models, Gini is a partially useful measure. This is unlike in macroeconomic models where exponential growth rate gives rise a appearance of non-trivial Gini coefficients}.  On the other hand, the mathematical form of expressions for $\mathcal{S}(t)$ suggest a clear and non-trivial dependence of the $D$ and clear separation between the population of particles that are isolates and the population of particles that are in a proto.

As the equations (\ref{e}) and (\ref{s}) show, all agents, irrespective of their starting intial condition and luck have a non-zero probability of dying. The paper that inspired this work ~\cite{srimil} emphasized studied inequality dynamics in the context of extreme scarcity in their foraging world. One can mimic this scarcity by simply turning of the income, leaving the agents to all inevitably die. The differential rate of death of the isolates and the non-isolates then become an important quantity.

In the simulation experiments that we discuss next discusses various stages and stopping conditions. All these stages and transition times between them make sense in terms of the underlying dynamics of the particle ensemble. In each of the stages, the effective dynamics are different from each other but the form of the equations remain the same.

To break the homogeneity of this particle ensemble, we propose to let the particles interact with only  a random subset of other populations. In other words, we embed the particles on a network (graph); particles interact (form proto) with only their neighbors. In this work, we focus on the simplest \textit{textbook} example of network: Erd\H{o}s-R\'{e}nyi network model~\cite{newman2018networks}. ER models has only one parameter $\lambda$, a Poisson parameter that determines the density of connections in the network, and in this setting, determines the number of particles available for a particle to coalesce and form protos. Commonsense reasoning suggests that a larger $\lambda$ parameter may help increase the average lifetimes of individual agents through by enabling greater likelihood of contact with other particles.

Just as equations (\ref{e}) and (\ref{s}) offers insight about the role of dynamics, the key feature of ER models is the the formation of giant component at $\lambda=1$, allowing a finite fraction of the population to be connected, have cliques of very high order, and other useful properties guarantee proto formation in large fraction of the population.

As its network properties are well-understood, wealth dynamics on such network allows us to understand more thoroughly, the interaction between dynamics and structure, and its role in global behavior. Still, a full analysis of this model remains to be completed and is part of a forthcoming work involving other more expressive network models. As we note subsequently, even this simple setting offers interesting insights,

@@

Various statistical counters are updated as the program executes so that its behavior can be analyzed post mortem. The main simulation loop can also be invoked from a ``parameter sweep'' program which executes it multiple times over a range of parameter values, in order to determine how the model's behavior changes in response to key parameters.

@@

\section{Analysis}
\input{analysis}
%%%%% need to add some speculations and motivate further simulation experiments .... especially what insights were gained by the current simulation model
\section{Conclusion and Future Work}
Being one of the first in an ongoing project, the next directions are too many to mention here. Still, we mention a few next steps that adheres to the spirit of our project. These directions all alleviate some of the obvious limitations of the current work.

While both the stochastic models and ER models are well understood, proto formation dynamics, as coagulation processes in continuous-time, remain to be better understood, mathematically. Also, while the mathematical analysis of dynamics of discrete-time stochastic models on networks are well studied, diffusion processes on networks seem under explored in the literature.

Even with only a few parameters, the simulation results are hard to visualize and interpret. As the number of parameters increase, as expected in future models, ideas from design and analysis of (computer) experiments may be required. Also, our exploration of the time to death distirbutions of the population show non-trivial stucture. More careful statisitcal tests that go beyond the mean analysis presented here are required to tease out the necessary and sufficient conditions for proto population to consistenty out-perform isolate population.

The models used for our simulation are at best \textit{stylized} models of real world social and economic systems, especially in anthropology~\cite{white2011kinship} and historical and urban sociology~\cite{sampson2012great,katz2013people}. Research in economic anthropology~\cite{koster2019,koster2014,koster2015,bogerhoff2015,smith2019,nolin2012,power2018cooperation,power2018} suggest complex food aand economic resource sharing rituals among members of various communities. Such resource sharing social networks do not look like any of the \textit{textbook} models. Extending our analysis to more expressive network models like exponential random graph models, stochastic block models and latent space models is an important research direction. This alongside the use of empirically observed cooperation and coordination protocols have potential in making our models better callibrated with real world systems.

Despite their simplicitly, models like the ones constructed here have several advantages.  As ACE models, they offer insights about economic systems in which majority of the assumptions of neo-classical economics like perfectly mixed agents~\cite{wilhite} and presence of equilibrium~\cite{arthur} do not hold. As AS models~\cite{hedstrom2011oxford}, they offer an approach that adds models of social mechanisms to CSS models in a graded manner.

>>>>

\maketitle
\begin{abstract}
We study wealth accumulation dynamics in a population of heterogeneously mixed agents with a capacity for a certain primitive form of cooperation enabled by static network structures. Despite their simplicity, the stochastic dynamics generate inequalities in wealth reminiscent of real-world social systems even in a fully mixed population. A simple form of cooperation is introduced and is shown to enhance the viability of agents by embedding such dynamics in a network; the impact of social structures on the origins and persistence of inequality can be teased out easily. The models developed here complement traditional modeling approaches based on grid worlds.

\end{abstract}
\section{Introduction}
In recent years, the concern and debate regarding wealth inequality and socioeconomic mobility have been one of the few unifying issues dominating the extremely polarized public spheres of the Global North. While economic and political inequality used to be discussed in heterodox economics circles, the contentious discussions on the topic within mainstream economics since the publication of Piketty's book~\cite{piketty2017capital} suggest a lack of consensus about basic foundational questions like the origins and persistence of economic inequality. Not surprisingly, traditional theories and tools of macro and micro economics are now being diagnosed for their limitations. Simultaneously, insights from related disciplines, along with novel models for scientific inquiry not usually associated with traditional econometrics, are being taken more seriously. The work presented here shares this spirit by integrating substantive ideas from anthropology, economic sociology~\cite{granovetter2017society} and urban sociology~\cite{sampson2012great} and using modeling approaches from computational social science~\cite{hedstrom2018}, computational economics~\cite{tesfatsion} and analytical sociology~\cite{hedstrom2011oxford} to explore the origins of inequality in a simple model of wealth dynamics in the presence of social structures.    
% SD: citation for "anthropology," the first element of the last sentence's list? Grief and/or White?

The current work originated in our attempts to incorporate and tease out the effects of social structures in simple models of wealth dynamics in the presence of environmental stochasticity, and a simple form of resource pooling. Specifically, we are interested in studying how structure and dynamics interact to produce and sustain inequality and to influence robustness to scarcity shocks. We answer the former question using Gini coefficients and the latter using survival time type analysis. As we discuss in section 3, while Gini was found to be of limited value, survival time analysis produced more insights on both questions of inequality and robustness.

These models were inspired by our search for analogs of the grid world agent-based models (ABMs) of Friesen and Mudigonda~\cite{srimil} where foraging agents that pool their resources were shown to on-average outperform non-pooling agents. The original model used foraging to mix the population and create opportunities for interaction, and, when certain conditions were met, resource sharing. Our model achieves agent interaction by postulating a static network structure which partially mixes the agents. Of primary interest is whether network effects alone can generate and sustain differences in economic outcomes of otherwise homogeneous agents.

The original model drew its inspiration from historical sociology, in Katz's influential study of middle of 19th century Hamilton, Canada~\cite{katz2013people}. Retaining this original motivation, we draw additional inspiration from economic sociology in the work of Granovetter~\cite{granovetter2005}; urban sociology, in the work of Sampson~\cite{sampson2002} and others; and in anthropology, in the work on cooperation in small to medium scale societies~\cite{avner1994,white2011kinship}, and others. The geographic and economic scale of the systems and the nature of social actors and time period of interest are all very different from the ones used to develop models of representative agents in macroeconomics~\cite{benhabib2018}, making the similarity between macroeconomic wealth dynamics models and our models not comparable without further justification. Elaborating on the interplay of conceptual and methodological ideas among these disciplines is beyond the scope of this article. Instead, we anchor this work in economic sociology and revisit the insights from the above disciplines in the article's concluding section.

The important role played by social structures or non-economic structures in determining economic outcomes of individuals in a society is not in doubt~\cite{granovetter2005, jackson_rev2017}. Still, in the absence of a unifying foundation for sociology and economics, the full impact of this two-way interpenetration of economic and social structures is demonstrated only on a case-by-case basis. The language of social and economic networks affords a first principle integration by simplifying the non-trivial concept of social and economic structure~\cite{martin_lee} to only dyadic relations.

As mentioned earlier, macroeconomic models are ill-suited to the study of collective phenomena at an intermediate level. Hence, alternative explanations of aggregate phenomena that match the expressiveness of economic models are required. Analytical sociology~\cite{ch1as_hdbk}, with its emphasis on explanation of collective emergent phenomena using mathematically formulated social mechanisms~\cite{ch2as_hdbk,ch11as_hdbk} anchored at the individual level, is an ideal candidate for this purpose. It is particularly powerful when combined with computational simulation, because when mathematically formulated models reach even a modest level of complexity, they often become analytically intractable. Simulation \textit{in silico} can yield approximate results for these more complex scenarios, which supplement the exact results reached by traditional methods.

@@

\label{giniKindaSucks}
% SD: In light of recent discussions, we may want to downplay Gini and up-play the robustness of agents to exogenous shocks.
The phenomenon under scrutiny is the emergence and evolution of economic inequality in societies with non-trivial social structures that enable economic interactions and cooperation mechanisms. The model we use to answer questions surrounding this phenomenon must possess a dynamic model of wealth accumulation, a model of social structure, and a suitably useful measure of inequality. The dynamics are modeled as Brownian-noise-driven linear dynamical systems, here with constant growth rate; the structure is modeled by networks, here with random graphs; the measures of inequality used are Gini coefficients, and survival time distributions of agents in response to lack of resources. Although both the dynamical system and the network model are quite well understood, the precise interplay of structure and dynamics produces interesting emergent wealth distributions and robustness to scarcity. To the best of our knowledge, this specific combination of network structure and social dynamics has not been discussed in the computational social science (CSS) or mathematical sociology literature, and we consider this the primary contribution of this work.

@@

In the next section, we discuss the mathematical formulation of the model. In section 3, we present the analysis of our simulation experiments, summarize key findings, and discuss why we chose the \texttt{julia} programming language~\cite{Julia-2017} for our implementation. In section 4, we discuss the limitations of our simple models, extensions to dynamics and networks more expressive than the ones presented here and planned future work.

Before discussing our model in greater mathematical detail, we discuss the model qualitatively, contrasting it with more familiar modeling approaches. The model presented here has much in common with models used in social-reality-inspired models in statistical physics~\cite{redner2001guide}, dynamic process models in network science~\cite{newman2018networks}, and computational social science and agent-based models~\cite{abm_review}; however, our modeling philosophy is somewhere at the interface of agent-based computational economics (ACE)~\cite{tesfatsion} and analytical sociology (AS)~\cite{ch1as_hdbk,ch2as_hdbk}. In ACE, we acknowledge its aspiration to develop bottom-up models of economic systems at all scales but restrain from its enthusiastic use of complex but well-calibrated detailed models of markets and agents~\cite{tesfatsion2017}. In AS, we assent its focus on social mechanisms in explaining social phenomena, but instead rely on simplified mechanisms with few parameters~\cite{ch11as_hdbk} with the specific goal of extracting insights from \textit{stylized} models.

More specifically, from statistical physics, we borrow the dynamics: diffusion models and associated first-passage time techniques; from network science, we borrow the structural aspects: Erd\H{o}s-R\'{e}nyi network (ER) models; and from non-equilibrium statistical physics, CSS and ACE, we borrow a form of cooperation: the concept of coalescence, institution and coordination.

\section{Model}
The model of cooperative wealth accumulation constructed here is best thought of as a stochastic interacting particle system infused with economic sociological semantics. The particle evolves according to a one dimensional diffusion process with constant drift and is driven by a Brownian noise with a boundary condition at the origin corresponding to particle absorption. After crossing a pre-determined threshold in state space, particles above the threshold follow a protocol and coalesce together. In an ensemble of otherwise identical particles, a given particle may coalesce with a subset of other particles. This mixing characteristic is encoded via a graph. Both the conjoined particles and individual particles die upon crossing the origin. This model of interacting diffusing particles can be provided with substantive semantics as follows.

%Might be nice to put a nice table here 
The one-dimensional state space of the particle is identified with the wealth of a social actor (agent). We consider a homogeneous population where all agents are required burn their wealth at a constant specified rate in order to survive. In addition, the agents all gain wealth at a constant rate. The resource draining rate and the resource gaining rate are additive and constitute the drift of the diffusion process. The environmental contingency is modeled by a Brownian noise of a specified intensity. Upon crossing a specific wealth threshold, agents can coalesce to form a cooperative unit deciding to pool their resources and their environmental contingencies into a single unit which we call a \textit{proto-institution} (proto). The mixing characteristic of this ensemble is the social network (here ER model).

The questions of interest to us, emergence and persistence of inequality and robustness to scarcity, in the presence of social structure and uncertain environmental conditions, map onto questions about the stochastic dynamical system (SDS). Inequality can be quantified using Gini coefficients of the particle ensemble's state space. Robustness to scarcity can be measured using survival time analysis for the particles to reach the absorbing boundary at the origin by turning off the appropriate drift terms in the dynamics.

@@

\noindent where $v$ is the wealth growth rate (the difference of the income and metabolic rate of the agent) and $D$ is the intensity of the Brownian process (white noise process) $w$. $x(t)$ is the state of the particle (wealth) at time $t$. The dynamics can be started at any initial point $x_0 > 0$.

@@

Since the particles are independent, they are independent dynamical systems; proto formation is a higher-level construct imposed on this system and is a constraint on what states are considered viable and which are not. Consider two particles $x_1$ and  $x_2$ that have crossed the threshold for coalescence $x_{thresh}$. After the coalescence event, the state (wealth) of individual particles $x_1(t)$ and  $x_2(t)$ is not important; only the aggregate wealth of the coalescent (proto) is important. As long as $x_1(t) + x_2(t) >0$, both particles survive. Effectively, the proto is a two-dimensional SDS. Since the wealth of the proto ($p_{12}$) is additive, the aggregate wealth $p_{12} = x_1(t) + x_2(t) $. The dynamical variable satisfies an SDE\footnote{The result follows from the additivity properties of white noise.}

@@

For such ensemble of particles, the primary driver of difference in paths (life histories) and wealth is the Brownian noise intensity $D$; greater $D$ leads to greater diversity (and hence Gini). However, Gini is a measure that is dependent on absolute magnitude. So, if the drift ($v$) is large, then the variation generated by
$D$ gets washed out by $v$\footnote{Preliminary investigations suggest that for simple non-network ensembles, Gini either stays close to $0$ or $1$. We suspect that this is because of the constant wealth growth rate used in our models, Gini is a partially useful measure. This is unlike in macroeconomic models where exponential growth rate gives rise an appearance of non-trivial Gini coefficients}.  On the other hand, the mathematical form of expressions for $\mathcal{S}(t)$ suggest a clear and non-trivial dependence of the $D$ and clear separation between the population of particles that are isolates and the population of particles that are in a proto.

As the equations (\ref{e}) and (\ref{s}) show, all agents, irrespective of their starting initial condition and luck, have a non-zero probability of dying. The paper that inspired this work ~\cite{srimil} studied inequality dynamics in the context of extreme scarcity in their foraging world. One can mimic this scarcity by simply turning off the income, leaving the agents to all inevitably die. The differential rate of death of the isolates and the non-isolates then becomes an important quantity.

In the simulation experiments that we discuss next discusses various stages and stopping conditions. All these stages and transition times between them make sense in terms of the underlying dynamics of the particle ensemble. In each of the stages, the effective dynamics are different from each other, but the form of the equations remains the same.

To break the homogeneity of this particle ensemble, we propose to let the particles interact with only a random subset of other populations. In other words, we embed the particles on a network (graph); particles interact (form proto) with only their neighbors. In this work, we focus on the simplest \textit{textbook} example of network: Erd\H{o}s-R\'{e}nyi network model~\cite{newman2018networks}. ER models have only one parameter $\lambda$, a Poisson parameter that determines the density of connections in the network, and which in this setting
determines the expected number of particles available for a particle to
coalesce with and form a proto. Commonsense reasoning suggests that a larger $\lambda$ parameter may help increase the average lifetimes of individual agents by enabling greater likelihood of contact with other particles.

Just as equations (\ref{e}) and (\ref{s}) offer insight into the role of dynamics, the key feature of ER models is the formation of giant component at $\lambda>=1$, allowing a finite fraction of the population to be connected, have cliques of very high order, and other useful properties guarantee proto formation in large fraction of the population.

As its network properties are well-understood, the network’s wealth dynamics allow us to understand more thoroughly the interaction between dynamics and structure, and its role in global behavior. Still, a full analysis of this model remains to be completed and is part of a forthcoming work involving other more expressive network models. As we note subsequently, even this simple setting offers interesting insights,

@@

Various statistical counters are updated as the program executes so that its behavior can be analyzed postmortem. The main simulation loop can also be invoked from a ``parameter sweep'' program which executes it multiple times over a range of parameter values, in order to determine how the model's behavior changes in response to key parameters.

@@

\section{Analysis}
\input{analysis}
%%%%% need to add some speculations and motivate further simulation experiments .... especially what insights were gained by the current simulation model
\section{Conclusion and Future Work}
This is one of the first in an ongoing project and the next directions are too many to mention here. Still, we mention a few next steps that adhere to the spirit of our project. These directions all alleviate some of the obvious limitations of the current work.

While both the stochastic models and ER models are well understood, proto formation dynamics, as coagulation processes in continuous time, remain to be better understood, mathematically. Also, while the mathematical analysis of dynamics of discrete-time stochastic models on networks is well studied, diffusion processes on networks seem under explored in the literature.

Even with only a few parameters, the simulation results are hard to visualize and interpret. As the number of parameters increases, as expected in future models, ideas from design and analysis of (computer) experiments may be required. Also, our exploration of the time to death distributions of the population show non-trivial structure. More careful statistical tests that go beyond the mean analysis presented here are required to tease out the necessary and sufficient conditions for
agents in protos (non-isolates) to consistently outperform the isolates.

The models used for our simulation are at best \textit{stylized} models of real world social and economic systems, especially in anthropology~\cite{white2011kinship} and historical and urban sociology~\cite{sampson2012great,katz2013people}. Research in economic anthropology~\cite{koster2019,koster2014,koster2015,bogerhoff2015,smith2019,nolin2012,power2018cooperation,power2018} suggests complex food and economic resource sharing rituals among members of various communities. Such resource sharing social networks do not look like any of the \textit{textbook} models. Extending our analysis to more expressive network models like exponential random graph models, stochastic block models and latent space models is an important research direction. This alongside the use of empirically observed cooperation and coordination protocols have potential in making our models better calibrated with real world systems.

Despite their simplicity, models like the ones constructed here have several advantages.  As ACE models, they offer insights about economic systems in which the majority of the assumptions of neo-classical economics like perfectly mixed agents~\cite{wilhite} and presence of equilibrium~\cite{arthur} do not hold. As AS models~\cite{hedstrom2011oxford}, they offer an approach that adds models of social mechanisms to CSS models in a graded manner.
